---
title: "MDA 9169A Project"
author: "Moustafa Shaker, Huicong(Ivy) Wu"
output: 
    html_document: 
       keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read Data
```{r}
#house price data
data = read.csv("kc_house_subset.csv")
#predict the house sales in King County, Washington State, USA
#consisted of historic data of houses sold between May 2014 to May 2015. 
#understand which factors are responsible for higher property value - $650K and above
```

1. Data Description
```{r}
library(dplyr)
dim(data)#dimension of dataset
str(data)#data type for each variables
```
2.Data information
```{r}
library(funModeling) 
library(tidyverse) 
library(Hmisc)
df <-data[,-c(1)]#exclude id
data_stats <- df_status(data)#check data unique, non-zeros, zeros, type and NA values.
arrange(data_stats, -p_zeros) %>% select(variable, q_zeros, p_zeros)# order data by percentages of zeros
```

Waterfront can be categorical variable.
```{r}
# high carinality variables
#one in which it can take many different values
arrange(data_stats, unique) %>%select(variable, unique)# frequency of lowest rank of variable
# frequency of continous variables
freq(df$bedrooms)
freq(df$grade)
freq(df$floors)
freq(df$view)
freq(df$condition)
freq(df$waterfront)
freq(df$bathrooms)
```

3.Data Prepocessing and Feature Construction
3.1 Handle variable"date" format
```{r}
df$date = substr(df$date, 1, 6)
# Converting it to numeric as we can only use numeric values for corrleation
df$date = as.numeric(as.character(df$date))
```

3.2 Add new feature age and renov_ag
```{r}
last_renew<-1:nrow(df)
index =1
df$start = substr(df$date, 1, 4)
df$age<-as.numeric(as.character(df$start))-df$yr_built
for (i in df$yr_renovated){
  #print(i)
  if (i != 0){
     # print(inedx)
      last_renew[index]=as.numeric(df$start)[index]-i
  }
  else{last_renew[index]=0}
  index=index+1
}
df$renov_age<-last_renew
#drop yr_built, y_renovated, start, date
drop <- c("date","start","yr_renovated","yr_built")
df<-df[!(names(df) %in% drop)]
```

3.3 Remove rows which bathroom ==0 and bedroom ==0
```{r}
df <-df[df$bathrooms > 0,]
df <-df[df$bedrooms > 0,]
```

3.4  add dummy variables for zipcode
```{r}
zip_ = factor(df$zipcode)
dummies = model.matrix(~zip_)
df <-cbind(df,dummies[,-c(1)])
#drop zipcode
drop <- c("zipcode")
df<-df[!(names(df) %in% drop)]
```

3.5 add new feature distance to center
```{r}
new<-paste(df$lat,sep=",",df$long)
distance <-1:nrow(df)
dist_to_center <- function (lat1, lon1, lat2, lon2)
{   
    lon2 <- lon2*pi/180;
    lat2 <- lat2*pi/180;
    lon1 <- lon1*pi/180;
    lat1 <- lat1*pi/180;
    dlon <- lon2 - lon1
    dlat<- lat2 - lat1
    a <- sin(dlat/2)^2 + cos(lat1) * cos(lat2) * sin(dlon/2)^2
    c <- 2 * atan2( sqrt(a), sqrt(1-a) ) 
    d <- 3961 * c;
} 
#print(dist_to_center(47.3487,-122.036,47.6062, 122.3321))      
for (i in 1:nrow(df)){
   distance[i]<-dist_to_center(df$long[i],df$lat[i], -122.200676,47.610378)*1.6
}
df$distance_to_center<-distance
# drop lat and long
drop <- c("lat","long")
df<-df[!(names(df) %in% drop)]
head(df)
```

3.6 add new feature bed to bath ratio
```{r}
bed_to_bath<-df$bedrooms/df$bathrooms
df<-cbind(df,bed_to_bath)
```

3.7 add new feature ratio between the square footage of the interior living space and the land lot
```{r}
living_to_lot<-df$sqft_living/df$sqft_lot
df<-cbind(df,living_to_lot)
```

3.8 add new feature the price per square footage
```{r}
price_per_sqrt<-df$price/df$sqft_lot
df<-cbind(df,price_per_sqrt)
```

3.9 add new feature the price per interior living space 
```{r}
price_per_interior<-df$price/df$sqft_living
df<-cbind(df,price_per_interior)
```

3.10 add new feature the price per floor
```{r}
price_per_floor<-df$price/df$floors
df<-cbind(df,price_per_floor)
```

3.11 add new feature Ratio between floors and land
```{r}
floors_to_land<-df$floors/df$sqft_lot
df<-cbind(df,floors_to_land)
```

3.12 add new feature Ratio of house living space to nearest 15 houses
```{r}
living_living15<-df$sqft_living/df$sqft_living15
df<-cbind(df,living_living15)
```

3.13 add new feature Ratio of land lot to nearest 15 houses
```{r}
lot_lot15<-df$sqft_lot/df$sqft_lot15
df<-cbind(df,lot_lot15)
```

3.14 add new feature bed to sqft_living 
```{r}
bed_living<-df$bedrooms/df$sqft_living
df<-cbind(df,bed_living)
```

3.15 add new feature price per bedroom 
```{r}
price_bed<-df$price/df$bedrooms
df<-cbind(df,price_bed)
```

3.16 add new feature bed to floor 
```{r}
bed_floor<-df$bedrooms/df$floors
df<-cbind(df,bed_floor)
```

3.17 add new feature bed to land
```{r}
bed_land<-df$bedrooms/df$sqft_lot
df<-cbind(df,bed_land)
```

3.18 add new feature price per bathrooms
```{r}
price_per_bathroom<-df$price/df$bathrooms
df<-cbind(df,price_per_bathroom)
```

3.19 add new feature price per condition 
```{r}
price_per_condition<-df$price/df$condition
df<-cbind(df,price_per_condition)
```

3.20 add new feature price per grade 
```{r}
price_per_grade<-df$price/df$grade
df<-cbind(df,price_per_grade)
```

3.21 add new feature price per sqft_above
```{r}
price_per_sqftabove<-df$price/df$sqft_above
df<-cbind(df,price_per_sqftabove)
```

3.22 add new feature price per distance from centre 
```{r}
price_per_dist<-df$price/df$distance_to_center
df<-cbind(df,price_per_dist)
```

3.23 add new feature bathroom to sqft_living
```{r}
bathroom_to_sqft_living<-df$bedrooms/df$sqft_living
df<-cbind(df,bathroom_to_sqft_living)
```

3.24 add new feature bathroom to floors 
```{r}
bathroom_to_floors<-df$bathrooms/df$floors
df<-cbind(df,bathroom_to_floors)
```

3.25 add new feature sqftliving_floots
```{r}
sqftliving_floors<-df$sqft_living/df$floors
df<-cbind(df,sqftliving_floors)
```

3.26 add new feature sqftliving_sqftabove
```{r}
sqftliving_sqftabove<-df$sqft_living/df$sqft_above
df<-cbind(df,sqftliving_sqftabove)
```

3.27 add new feature age_dif_renovate
```{r}
age_dif_renovate<-df$age-df$renov_age
df<-cbind(df,age_dif_renovate)
```


4. Exploratory Data Analysis
4.1 Pairplots among all variables
```{r}
pairs(price~bedrooms+bathrooms+sqft_living+floors,data = df,col = "blue",pch = 16)
pairs(price~sqft_lot+sqft_above+sqft_basement+sqft_living+sqft_living15+sqft_lot15,data = df,col = "blue",pch = 16)
pairs(price~waterfront+condition+view+grade,data = df,col = "blue",pch = 16)#categorical variables
```

4.2 Analyze response variable - house price distribution
```{r}
#install.packages("ggpubr")
library(ggpubr)
# Basic density plot with mean line and marginal rug
a <- ggplot(df, aes(x = price))
a + geom_histogram(aes(y = ..density..), 
                   colour="black", fill="white",bins=30) +
  geom_density(alpha = 0.2, fill = "#FF6666") 
boxplot(df$price)
```

4.3 Analyze how age and yr_renew of houses vs. price
```{r}
plot(df$price~df$age)
plot(df$price~df$renov_age)
```

4.4 Find bathrooms relation with house price
```{r}
price_bath<- select(df,price,bathrooms)
boxplot(price_bath$price~price_bath$bathrooms,xlab="bathrooms",ylab="House Price",main="bathrooms vs. house price",col = "blue")
```

4.5 Find how sqft_living related to house price
```{r}
price_area<- select(df,price,sqft_living)
plot(price_area$price~price_area$sqft_living,xlab="sqft_living",ylab="House Price",main="sqft_living vs. house price",pch =16,col ="blue")
```

4.6 Find how sqft_above related to house price
```{r}
price_above<- select(df,price,sqft_above)
plot(price_above$price~price_above$sqft_above,xlab="sqft_above",ylab="House Price",main="sqft_above vs. house price",pch=16,col ="blue")
```

4.7 Find how sqft_basement related to house price
```{r}
price_above<- select(df,price,sqft_basement)
plot(price_above$price~price_above$sqft_basement,xlab="sqft_basement",ylab="House Price",main="sqft_basement vs. house price",pch=16,col ="blue")
```

4.8 Find how sqft_living15 related to house price
```{r}
price_sqft15<- select(df,price,sqft_living15)
plot(price_sqft15$price~price_sqft15$sqft_living15,xlab="sqft_living",ylab="House Price",main="sqft_living15 vs. house price",pch=16,col ="blue")
```

4.9 lat and long to center
```{r}
plot(df$price~df$distance_to_center)
#min(df$distance_to_center)
#max(df$distance_to_center)
#df
```

4.10 Find how grade affects the house price
```{r}
price_grade<- select(df,price,grade)
boxplot(price_grade$price~price_grade$grade,xlab="grade",ylab="House Price",main="grade vs. house price",col =c("yellow","lightyellow","blue","red","green","orange","purple","pink","grey") ,border = "brown")
abline(h =mean(price_grade$price),col="red",lwd=3, lty=2)
plot(price_grade$price~price_grade$grade,xlab="grade",ylab="House Price",main=" Scatterplot of grade vs. house price")
```

4.11 Find relationship with View and price
```{r}
price_view<- select(df,price,view)
boxplot(price_view$price~price_view$view,xlab="view",ylab="House Price",main="view vs. house price",col = "blue")
plot(price_view$price~price_view$view,xlab="view",ylab="House Price",main="Scatterplot of view vs. house price")
```

4.12 Find relationship with condition and price
```{r}
price_condition<- select(df,price,condition)
boxplot(price_condition$price~price_condition$condition,xlab="condition",ylab="House Price",main="condition vs. house price",col = "blue")
plot(price_condition$price~price_condition$condition,xlab="condition",ylab="House Price",main=" Scatterplot of condition vs. house price")
```
4.13 Find relationship with waterfront and price
```{r}
price_waterfront<- select(df,price,waterfront)
boxplot(price_waterfront$price~price_waterfront$waterfront,xlab="waterfront",ylab="House Price",main="waterfront vs. house price",col = "blue")
```
4.14 Find relationship with floors and price
```{r}
price_floors<- select(df,price,floors)
boxplot(price_floors$price~price_floors$floors,xlab="floors",ylab="House Price",main="floors vs. house price",pch =16,col ="blue")
```

4.15 Find relationship with price and bed_to_bath
```{r}
plot(df$price~df$bed_to_bath)
```
4.16 Find relationship with price and living_to_lot
```{r}
plot(df$price~df$living_to_lot)
```
4.17 Find relationship with price and price/sqrt,price/interior,price_per_floor, price/sqft_above,price_per_dist
```{r}
plot(df$price~df$price_per_sqrt)
plot(df$price~df$price_per_interior)
plot(df$price~df$price_per_floor)
plot(df$price~df$price_per_sqftabove)
plot(df$price~df$price_per_dist)
```

4.18 Find relationship with price and price/bed,price/bathroom,price/condtion,price/grade, 
```{r}
plot(df$price~df$price_per_bathroom)
plot(df$price~df$price_bed)
plot(df$price~df$price_per_condition)
plot(df$price~df$price_per_grade)
```

4.19 Find relationship with price and floors_to_land,living_living15,lot_lot15,bed_living,bed_land,bed_floor,bathroom_to_floors,sqftliving_floots,sqftliving_sqftabove
```{r}
plot(df$price~df$floors_to_land)
plot(df$price~df$living_living15)
plot(df$price~df$lot_lot15)
plot(df$price~df$bed_living)
plot(df$price~df$bed_land)
plot(df$price~df$bed_floor)
plot(df$price~df$bathroom_to_floors)
plot(df$price~df$sqftliving_floors)
plot(df$price~df$sqftliving_sqftabove)
plot(df$price~df$age_dif_renovate)
```

```{r}
df_new<-df[-c(17:85)]#drop zipcode...
#df_new
#information about continuous variables, including quantiles, mean, median, count, missing values.
exclude_ <-c("waterfront")
exclude_df<-df[ , !(names(df) %in% exclude_)]
describe(exclude_df)
```
```{r}
ncol(df_new)
```

4.19 Correlation Matrix and Correlation Heatmap
```{r}
library(ggplot2)
library(reshape2)
#rebuild the dataframe
corr_matrix_data <- cor(df_new)
#plot correlation matrix
head(corr_matrix_data)#linear dependence between two variables
print("rank of correlation index")
sort(corr_matrix_data[,1])
#default is "pearson"
#plot heatmap
melted_corr_matrix <- melt(corr_matrix_data)
ggplot(data = melted_corr_matrix, aes(x=Var1, y=Var2, fill=value)) + geom_tile() +
theme(text = element_text(size=12),axis.text.x = element_text(angle=90, hjust=1),plot.title = element_text(hjust=0.5))+ggtitle("Correlation Heatmap")+labs( fill = expression("Relation"))+scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1),name="Pearson\nCorrelation") +geom_text(aes(Var2, Var1, label = round(value,2)), color = "black", size = 2)
plot_num(df_new)
```


4.20 print  correlation values higher than 0.5
```{r}
zdf <- as.data.frame(as.table(cor(df_new)))
subset(zdf, abs(Freq) > 0.5)
```

5. Model Selection
```{r}
#sum(is.na(df_new))
#df_new <- na.omit(df_new)
X =model.matrix(price~.,df_new)[,-1]
nrow(X)
y =df_new$price
length(y)
set.seed (10) 
n = nrow(df_new)
n
a<-sample(n ,round(0.6*n) , rep=FALSE) 
a
y_tr <- y[a]
X_tr <- X[a,]
y_ts <- y[-a]
X_ts <- X[-a,]
# fit linear model
fit_model = lm(price~., data =df_new[a,] )
summary(fit_model)
pred_model = predict(fit_model, newdata=df_new[-a,]) # prediction for test data
mse_model = mean((pred_model-y_ts)^2) # mse for test data
sqrt(mse_model)
```

OLS Regression with CV
```{r}
#OLS Regression with cross validation
#drop sqft_above
exclude_ <-c("sqft_basement")
df_new_ols<-df_new[ , !(names(df_new)%in%exclude_)]
n = nrow(df_new)
k = 20
RMSE_kcv_ols = numeric(k)
#Create k equally size folds
folds <- cut(1:n,breaks=k,labels=FALSE)
folds
#Perform a k-fold cross validation
for(i in 1:k)
{
  # Find the indices for test data
  test_index = which(folds==i)
  
  # Obtain training/test data
  test_data = df_new_ols[test_index, ]
  training_data = df_new_ols[-test_index, ]
  
  #Training matrix
  X_tr = model.matrix(price~.,training_data)[, -1]
  y_tr = training_data$price
  
  #Test matrix
  X_ts = model.matrix(price~.,test_data)[, -1]
  y_ts = test_data$price
  
  fit_ols = lm(price~., data =df_new[a,] )
 
  # Obtain RMSE on the 'test' data
  pred_ols = predict(fit_ols,newx = X_ts)
  resid_ols = y_ts - pred_ols 
  RMSE_kcv_ols[i] = sqrt(sum(resid_ols^2)/length(y_ts)) 
}
#All the values for RMSE
RMSE_kcv_ols
# Mean of RMSE 
mean(RMSE_kcv_ols)
```

Lasso Regression
```{r}
library(glmnet)
#LASSO with cross validation 
n = nrow(df_new)
k = 20
RMSE_kcv_lasso = numeric(k)
#Create k equally size folds
folds <- cut(1:n,breaks=k,labels=FALSE)
folds
#Perform a k-fold cross validation
for(i in 1:k)
{
  # Find the indices for test data
  test_index = which(folds==i)
  
  # Obtain training/test data
  test_data = df_new[test_index, ]
  training_data = df_new[-test_index, ]
  
  #Training matrix
  X_tr = model.matrix(price~.,training_data)[, -1]
  y_tr = training_data$price
  
  #Test matrix
  X_ts = model.matrix(price~.,test_data)[, -1]
  y_ts = test_data$price
  
  fit_lasso_cv = cv.glmnet(X_tr, y_tr, alpha = 1)
  bestlam = fit_lasso_cv$lambda.min
  fit_lasso_best = glmnet(X_tr, y_tr, alpha = 1, lambda = bestlam)
  
  # Obtain RMSE on the 'test' data
  pred_lasso = predict(fit_lasso_best, s = bestlam, newx = X_ts)
  resid_lasso = y_ts - pred_lasso 
  RMSE_kcv_lasso[i] = sqrt(sum(resid_lasso^2)/length(y_ts)) 
}
#All the values for RMSE
RMSE_kcv_lasso
# Mean of RMSE 
mean(RMSE_kcv_lasso)
```

Ridge Regression
```{r}
#Ridge regression with cross validation
n = nrow(df_new)
k = 20
RMSE_kcv_ridge = numeric(k)
#Create k equally size folds
folds <- cut(1:n,breaks=k,labels=FALSE)
folds
#Perform a k-fold cross validation
for(i in 1:k)
{
  # Find the indices for test data
  test_index = which(folds==i)
  
  # Obtain training/test data
  test_data = df_new[test_index, ]
  training_data = df_new[-test_index, ]
  
  #Training matrix
  X_tr = model.matrix(price~.,training_data)[, -1]
  y_tr = training_data$price
  
  #Test matrix
  X_ts = model.matrix(price~.,test_data)[, -1]
  y_ts = test_data$price
  
  fit_ridge_cv = cv.glmnet(X_tr, y_tr, alpha = 0)
  bestlam = fit_ridge_cv$lambda.min
  fit_ridge_best = glmnet(X_tr, y_tr, alpha = 0, lambda = bestlam)
  
  
  # Obtain RMSE on the 'test' data
  pred_ridge = predict(fit_ridge_best, s = bestlam, newx = X_ts)
  resid_ridge = y_ts - pred_ridge 
  RMSE_kcv_ridge[i] = sqrt(sum(resid_ridge^2)/length(y_ts)) 
}
#All the values for RMSE
RMSE_kcv_ridge
# Mean of RMSE 
mean(RMSE_kcv_ridge)
```

```{r}
coef(fit_ridge_best)
coef(fit_lasso_best)
```

Fit Lasso Regression with orginal variables without any feature engineering
```{r}
data <-data[data$bathrooms > 0,]
data <-data[data$bedrooms > 0,]
data_new<-data[-c(1:2)]#drop zipcode...
n = nrow(data_new)
k = 2
RMSE_kcv_lasso = numeric(k)
#Create k equally size folds
folds <- cut(1:n,breaks=k,labels=FALSE)
folds
#Perform a k-fold cross validation
for(i in 1:k)
{
  # Find the indices for test data
  test_index = which(folds==i)
  
  # Obtain training/test data
  test_data = data_new[test_index, ]
  training_data = data_new[-test_index, ]
  
  #Training matrix
  X_tr = model.matrix(price~.,training_data)[, -1]
  y_tr = training_data$price
  
  #Test matrix
  X_ts = model.matrix(price~.,test_data)[, -1]
  y_ts = test_data$price
  
  fit_lasso_cv = cv.glmnet(X_tr, y_tr, alpha = 1)
  bestlam = fit_lasso_cv$lambda.min
  fit_lasso_best = glmnet(X_tr, y_tr, alpha = 1, lambda = bestlam)
  
  # Obtain RMSE on the 'test' data
  pred_lasso = predict(fit_lasso_best, s = bestlam, newx = X_ts)
  resid_lasso = y_ts - pred_lasso 
  RMSE_kcv_lasso[i] = sqrt(sum(resid_lasso^2)/length(y_ts)) 
}
#All the values for RMSE
RMSE_kcv_lasso
# Mean of RMSE 
mean(RMSE_kcv_lasso)
```

```{r}
coef(fit_lasso_best)
```